Towards a standardized evaluation of imputation methodology

### 2022-03-31 Formatted/Formatted.tex
Process last feedback Gerko 31-03

\begin{document}
%\DOIsuffix{bimj.DOIsuffix}
\DOIsuffix{bimj.200100000}
\Volume{52}
\Issue{61}
\Year{2010}
\pagespan{1}{}
\keywords{Evaluation; Imputation; Missing data; Simulation studies;\\ [1pc]
% \noindent \hspace*{-4.2pc} Supporting Information for this article is available from Github, \break \hspace*{-4pc} \underline{github.com/gerkovink/StandardizedEvaluation}.
% } %%% semicolon and fullpoint added here for keyword style
\noindent \hspace*{-4pc} Supporting Information for this article is available online from \\
\hspace*{-4pc} \underline{github.com/gerkovink/StandardizedEvaluation}. 
}  %%% semicolon and fullpoint added here for keyword style

\title[Standardized evaluation]{Towards a standardized evaluation of imputation methodology}
%% Information for the first author.
\author[Oberman]{Hanne I. Oberman\footnote{Corresponding author: {\sf{e-mail: h.i.oberman@uu.nl}}}\inst{,1}} 
\address[\inst{1}]{Departement of Methodology \& Statistics, Padualaan 14, 3584 CH Utrecht, The Netherlands}
%%%%  Information for the second author
\author[Vink]{Gerko Vink\inst{1}}
%%%%  \dedicatory{This is a dedicatory.}
\Receiveddate{zzz} \Reviseddate{zzz} \Accepteddate{zzz}

\begin{abstract}
Developing new imputation methodology has become a very active field. Unfortunately, there is no consensus on how to perform simulation studies to evaluate the properties of imputation methods. In this paper we propose a move towards a standardized evaluation of imputation methods. To demonstrate the need for standardization, we highlight a set of potential pitfalls that bring forth a chain of potential problems in the objective assessment of the performance of imputation routines. This may lead to sub-optimal use of multiple imputation in practice. Additionally, we suggest a course of action for simulating and evaluating missing data problems.
\end{abstract}

@@

Imputation is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policymaking, demonstrated e.g. by the detailed manual created by the National Research Council \citep*{little2012prevention}. Although a top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers and data scientists are embracing imputation techniques. After all, the principle of imputation is very intuitive.

@@

When evaluating the statistical properties (and thereby the practical applicability) of imputation methodology, researchers most often make use of simulation studies. In such studies a complete dataset is usually generated from a statistical model, another model is used to induce missingness, and a set of evaluation criteria is postulated to evaluate the performance of one or more missing data methods. However, no golden standard has been established to evaluate imputation routines and, as a result, the validity and aim of the simulation set-ups may differ tremendously from one developer to another. Especially with novel imputation methods being propagated from the fields of machine learning and artificial intelligence, the differences may become more pronounced. Although these promising new methods seem to yield even sharper imputations than now-standard (semi-)parametric imputation methods, the comparison may not be fair due to the simulation setup.

% [Nowadays there are a lot of imp alg being propagated, esp ai/ml, whose perf often times seems quite promising. But may not be a fair comparison because of the simulation set-up/dev in ai/ml so fast that steps may be skipped. Especially in the field of ML/AI, there are promising new imputation methods, which may yield even sharper imputations than now-standard (semi-)parametric imputation methods]

@@

To evaluate the ability of an imputation routine to handle missingness, a form of truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand by choosing a data generating mechanism. Data generating mechanisms define how a complete dataset is obtained at the start of each simulation repetition. There are two general approaches to generating complete data: (i) model-based simulation, in which data are drawn from a known statistical model or probability distribution, such as the multivariate normal distribution; and (ii) design-based simulation, where data are sampled (with replacement) from a sufficiently large observed set, such as from official registers.

The problem with model-based data generating mechanisms is that a method's performance on simulated data may not translate to empirical data. Real-life data hardly ever follow a given theoretical distribution, so there is no guarantee that simulation results are generalizable. Moreover, data are often generated such that the problem that is being studied is most pronounced, e.g. with consistently high correlations between groups of variables. This results in simulated data that contain such valuable information structures that, no matter what type of missingness would subsequently be induced, the observed parts of the data will still hold much (if not all) of the information about the missing part. Unsurprisingly, the performance of any imputation method will then be evaluated as good. Another threat to the generalizability of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other (unfair) comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the variable type(s), and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit simulation conditions, which may give a false sense of objectivity.

@@

After deciding upon a data generating mechanism, another influential--but often overlooked--decision awaits. Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations, and only use the missing data generation procedure as source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, it is not necessary to take sampling variation into account in order to obtain information about a method's ability to handle the missing data problem, or to objectively compare methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation set-ups is to remove sampling variance from performance evaluation. A lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results.

@@

Evaluating imputation methodology requires a missing data problem to be solved. After establishing a comparative truth from a data generating mechanism, some form of missingness therefore has to be induced. Often, however, reports of simulation studies remain vague about the actual missingness conditions under investigation and, even worse, some authors only report something like:
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness mechanisms and missingness patterns. Missingness mechanisms describe the relationship between missing entries and observed data values and are generally categorized into MCAR, MAR and MNAR mechanisms \citep[][see Table \ref{table:mech}]{rubi76}. [TODO: maybe remove mechanisms here?] Missingness patterns concern the location of missing entries across incomplete data \citep[][p. 8]{litt20}. Under this definition, there is a row-wise element to the missingness pattern describing which variables are jointly observed, and a column-wise element encompassing the amount of missingness in the data.

Even the terminology on missingness generation can be confusing. Does a missingness proportion of 50\% mean that half of the entries in an incomplete dataset are missing, or that half of the rows have at least one missing entry? In this paper, we will henceforth refer to the latter as the proportion of incomplete cases, and keep the term `missingness proportion' restricted to the variable-by-variable interpretation. The distinction between the two concepts, however, is not always clear in the literature, and convoluting the terms may lead to incorrect generalizations because they rarely mean the same (e.g., a proportion of 50\% incomplete cases in bivariate data could translate to a missingness proportion of 25\% in both variables, or one completely observed variable and one with 50\% missingness). Moreover, the value of the actual missingness proportion may be diffusing too. Some studies use only 10\% missingness where other studies push the limits to additionally investigate the performance under missingness proportions of at least 50\%. The inconsistent display of simulation conditions may impact the objectivity of meta-evaluations over imputation methods, as one method's performance may appear to be favorable because of less stringent simulation conditions. This ultimately may lead to statisticians recommending a less efficient method to applied researchers, thereby limiting the efficiency of the imputation approach and unnecessarily lowering the statistical power.

@@

Another aspect of missingness patterns that may misguide evaluations is the complexity of the patterns across variables. It is hardly reasonable to imagine empirical data with only one incomplete variable, yet some simulation studies rely on univariate missingness patterns anyways. Extrapolating simulation results from a specific missing data pattern to more intricate empirical missingness patterns could lead to sub-optimal advice in practice. For example, if a simulator induces missingness in the outcome variable of their analysis model exclusively, they may conclude that list-wise deletion outperforms their imputation method(s). Such a conclusion would, unfortunately, only translate to data that adhere to the same special case \citep[described in][\S 2.7]{fimd}, while biasing inferences in other cases. Another example of a missingness pattern that may inadvertently impose assumptions on generalizations is a monotone pattern \citep[i.e. a pattern with uniformly increasing missingness proportions along variables;][]{litt20}. If a simulator investigates iterative imputation methods under monotone missingness, they might believe that the evaluated imputation algorithms converge instantly, while these methods actually require iteration to produce valid results in all other situations. A final example is multivariate missingness that is generated using step-wise univariate missingness induction. The resulting incomplete data may then not have the desired statistical properties \citep[e.g. skewness;][]{ampute}. This complicates any definitive conclusions about imputation method performance in relation to the data generating mechanism.

% is the   [Other problems with missingness patterns: monotone/univariate vs multivariate, sporadic vs systematic, and univariately generated missing data may not have the expected multivariate patterns.] For example, iterative imputation algorithms converge instantly if there is only one variable with missingness or when the pattern is monotone, but require iteration in all other situations. And very specific missing data patterns may misguide which imputation method prevails in performance: some ad hoc methods are known to yield valid inferences under certain restricted conditions \citep[e.g., list-wise deletion may outperform imputation methods if there is missingness in the outcome variable of the analysis model exclusively][]{fimd}, while biasing inferences in other cases. 
% [Missingness mechanisms paragraphs: 1) definitions of mechanisms (now in table), 2) that MCAR is often ignored, while it's the minimum requirement and reference, and sometimes realistic in practice, 3) that MAR can be generated \textit{spuriously} if the correlation between variables is low,  4) that MNAR is often ignored, while it may be the most realistic in practice, and 5) finally add something about types of M(N)AR (ref Rianne's Dance paper).]
% MCAR missingness: The scenario where the missing values are missing at random and the observed values are observed at random. 
%  the validity of assumptions become increasingly important when the missingness increases.

\begin{table}[tb]
\begin{center}
\caption{Missingness mechanisms. [TODO: add PMAR?]}
\label{table:mech}
\begin{tabular}{lll}
\hline
Mech. & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & Missing completely at random: The probability to be missing is the same for all \\
      & cases. In other words, the missing values are missing at random and the observed \\
      & values are observed at random. \\
MAR   & Missing at random: The probability to be missing is the same within groups of \\
      & cases defined by the observed data only. The essence is that observed relations in \\
      & the data inform the missingness, such that MCAR may be assumed within the \\ 
      & observed groups. \\
MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
      & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
      & inferred from the observed data. This is considered non-ignorable missingness \\
      & \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

Missingness mechanisms are usually assumed to be random (MAR; see Table \ref{table:mech} for definitions). However, the missingness mechanisms that are induced in simulation studies do not necessarily match any true missingness generating mechanisms in incomplete empirical data, nor the mechanisms that are typically assumed when imputing such data. On the one hand, there is MCAR missingness, which may be unlikely as the true missingness generating mechanism, but would yield insightful evaluations in a simulation setting. On the other hand, there is MNAR, which is arguably a quite reasonable source of many real-world missingness scenarios, but is generally ignored except for those evaluations that are specifically targeted at non-ignorable applications. A disconnect between induced missingness, real missingness, and assumed missingness may result in simulation studies that are not as informative as they could be.

@@

Although MAR missingness is often considered as simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to emulate a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is random. In fact, this would amount to one of the special cases under which complete case analysis would be more efficient than multiple imputation \citep[see e.g.][p. 48]{fimd}: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.

@@

The evaluation criteria used to assess imputation performance vary from one simulator to another. This is not surprising as people from different fields could have a different focus on the problem at hand. There are, however, some over-arching issues with assessing imputation method performance. In the first place, evaluations of the imputation-generating process (e.g., an iterative imputation algorithm) are generally left out of simulation study results. In the second place, there are performance measures pitfalls to consider.

Many contemporary imputation techniques rely on iterative algorithms, such as the Gibbs sampler, to generate imputations. As with any iterative algorithm--but especially with imputation algorithms that are critically considered to be possibly incompatible Gibbs samplers \citep[PIGS,][]{li2012imputing}--algorithmic convergence should be carefully evaluated. Unfortunately, there is no universal quantitative method to diagnose non-convergence in iterative imputation algorithms \citep{zhu15, ober21} and the alternative \citep[visual inspection of the imputation algorithm;][]{fimd} is neither efficient nor failproof. As a result, imputation algorithms may be terminated before reaching a stable state, which could yield sub-optimal imputations and under-estimated performance of the method. Moreover, failing to produce stable imputations is not an exclusive quality iterative aimputation algorithms. For example if a method does not yield any imputations at all in a certain simulation repetition (e.g. due to over-parameterization), this can be seen as failure in the imputation-generating process too.

The evaluation of simulated results relies on the relation between missing data models, imputation models, and analysis models (i.e. the complete data model used to estimate a target or estimand). When an imputation model is able to capture the essence of the true non-response mechanism relative to the analysis model, the models are said to be congenial \citep{meng94}. Congeniality is difficult to diagnose, but ignoring it may yield sub-optimal imputations. Methods to asses the suitability of imputation models and diagnose misfit often rely on visual inspection of the imputations \citep[see e.g.][]{abayomi2008diagnostics, bond16}, which is an infeasible endeavor in many simulation studies. Therefore, the performance of imputation procedures on distributional properties is often ignored in simulation studies. Even though the estimates on the analysis level may be justified, some methods can yield imputations that may seem completely invalid to applied researchers. For example, one could very accurately estimate average human height by filling in negative values and values that are unrealistically large. While the obtained inference could still be valid under such imputations, the plausibility of the imputed values given the observed data should be under scrutiny.

If the goal is inference--not prediction--then the uncertainty about estimates needs to be properly quantified. A single imputation would then not suffice. The aim of multiple imputation is not to reproduce the data, but to allow for obtaining valid inference given that the data are incomplete. This means that, given the framework provided by \citet{rubi87}, statistical properties such as bias, confidence intervals, and the coverage rate of the confidence intervals should be studied. After all, the 95\% confidence interval should contain the `true' value at least 95 out of 100 times \citep[][p. 591]{neym34}. Merely focusing on the discrepancy between the true and imputed data would open the door for invalid inferences.

Some simulators may focus on retrieving the unobserved values, while others study the performance in the context of obtaining efficient predictions or classification. For many outcome-driven imputation task a single valid imputation may perhaps suffice. We would like to note, however, that using measures of predictive accuracy or the RMSE (root mean squared error) of predicted values as the evaluation criterion may increase the rate of false positives \citep[][\S 2.6]{fimd}.

@@

\begin{table}[tb]
\begin{center}
\caption{Steps to consider in imputation simulation studies.}
\label{table:steps}
\begin{tabular}{lll}
\hline
&Step                    & Simulation aspects under consideration \\
\hline  
1&Set scope            & Aim(s), missing data method(s) to evaluate, simulation design \\
2&Obtain truth         & Data generating mechanism(s), estimand(s), sampling variance \\
3&Induce missingness   & Missingness mechanism(s), missing data pattern(s) \\
4&Apply methods        & Imputation model(s), analysis model(s) \\
5&Evaluate imputations & Algorithmic convergence, distributional characteristics, imputation \\
 &                      & model fit \\
6&Evaluate performance & Performance measures, qualitative judgement \\
7&Report               & Text, visualization, checklist, simulation script \\
\hline
\end{tabular}
\end{center}
% \footnotesize{Note. Multiple aspects may be varied across simulation conditions (e.g., using a factorial design with several data generating mechanisms and missingness proportions).}
\end{table}

@@

Before setting up a simulation study, the simulator should clearly define the scope of their evaluations. A simulation study aimed at imputation method comparison will typically have a different design than one aimed at establishing the inferential validity of a single (novel) method. The choice of imputation method(s) under evaluation is naturally intertwined with the simulation aims and set-up.

@@

Some general choices in the simulation set-up to consider at this stage are the simulation design and the number of simulation repetitions. The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There are, for example, known interaction effects between missingness mechanisms and missingness patterns (the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions). If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons. The number of simulation repetitions may be informed by the required level of precision in the simulation study \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error][]{morr18}.

@@

The simulator should choose their data generating mechanism(s) in line with the study scope. Model-based data-generating mechanisms have advantages in flexibility and precision, since data are generated from a known statistical model and the true theoretical parameters can be derived. A design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest.

>>>>

\begin{document}
%\DOIsuffix{bimj.DOIsuffix}
\DOIsuffix{bimj.200100000}
\Volume{52}
\Issue{61}
\Year{2010}
\pagespan{1}{}
\keywords{Evaluation; Imputation; Missing data; Simulation studies;\\ [1pc]
% \noindent \hspace*{-4.2pc} Supporting Information for this article is available from Github, \break \hspace*{-4pc} \underline{github.com/gerkovink/StandardizedEvaluation}.
% } %%% semicolon and fullpoint added here for keyword style
\noindent \hspace*{-4pc} Supporting Information for this article is available online from \\
\hspace*{-4pc} \underline{gerkovink.com/evaluation}. 
}  %%% semicolon and fullpoint added here for keyword style

\title[Towards a standardized evaluation of imputation methodology]{Towards a standardized evaluation of imputation methodology}
%% Information for the first author.
\author[Oberman]{Hanne I. Oberman\footnote{Corresponding author: {\sf{e-mail: h.i.oberman@uu.nl}}}\inst{,1}} 
\address[\inst{1}]{Departement of Methodology \& Statistics, Padualaan 14, 3584 CH Utrecht, The Netherlands}
%%%%  Information for the second author
\author[Vink]{Gerko Vink\inst{1}}
%%%%  \dedicatory{This is a dedicatory.}
\Receiveddate{zzz} \Reviseddate{zzz} \Accepteddate{zzz}

\begin{abstract}
Developing new imputation methodology has become a very active field. Unfortunately, there is no consensus on how to perform simulation studies to evaluate the properties of imputation methods. In this paper, we propose a move towards a standardized evaluation of imputation methods. To demonstrate the need for standardization, we highlight a set of potential pitfalls that bring forth a chain of potential problems in the objective assessment of the performance of imputation routines. This may lead to sub-optimal use of multiple imputation in practice. Additionally, we suggest a course of action for simulating and evaluating missing data problems.
\end{abstract}

@@

Imputation is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policymaking, demonstrated e.g. by the detailed manual created by the National Research Council \citep*{little2012prevention}. Although top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers and data scientists are embracing imputation techniques. After all, the principle of imputation is very intuitive.

@@

When evaluating the statistical properties (and thereby the practical applicability) of imputation methodology, researchers most often make use of simulation studies. In such studies a complete dataset is usually generated from a statistical model, another model is used to induce missingness, and a set of evaluation criteria is postulated to evaluate the performance of one or more missing data methods. However, no golden standard has been established to evaluate imputation routines and, as a result, the validity and aim of the simulation setups may differ tremendously from one developer to another. Especially with novel imputation methods being propagated from the fields of machine learning and artificial intelligence, the differences may become more pronounced. Although these promising new methods seem to yield even sharper imputations than now-standard (semi-)parametric imputation methods, the comparison may not be fair due to the simulation setup.

% [Nowadays there are a lot of imp alg being propagated, esp ai/ml, whose perf often times seems quite promising. But may not be a fair comparison because of the simulation setup/dev in ai/ml so fast that steps may be skipped. Especially in the field of ML/AI, there are promising new imputation methods, which may yield even sharper imputations than now-standard (semi-)parametric imputation methods]

@@

To evaluate the ability of an imputation routine to handle missingness, a form of truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand by choosing a data-generating mechanism. Data-generating mechanisms define how a complete dataset is obtained at the start of each simulation repetition. There are two general approaches to generating complete data: (i) model-based simulation, in which data are drawn from a known statistical model or probability distribution, such as the multivariate normal distribution; and (ii) design-based simulation, where data are sampled (with replacement) from a sufficiently large observed set, such as from official registers.

The problem with model-based data-generating mechanisms is that a method's performance on simulated data may not translate to empirical data. Real-life data hardly ever follow a given theoretical distribution, so there is no guarantee that simulation results are generalizable. Moreover, data are often generated such that the problem that is being studied is most pronounced, e.g. with consistently high correlations between groups of variables. This results in simulated data that contain such valuable information structures that, no matter what type of missingness would subsequently be induced, the observed parts of the data will still hold much (if not all) of the information about the missing part. Unsurprisingly, the performance of any imputation method will then be evaluated as good. Another threat to the generalizability of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other (unfair) comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the variable type(s), and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit simulation conditions, which may give a false sense of objectivity.

@@

After deciding upon a data-generating mechanism, another influential--but often overlooked--decision awaits. Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations and only use the missing data generation procedure as the source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, it is not necessary to take sampling variation into account in order to obtain information about a method's ability to handle the missing data problem, or to objectively compare methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation setups is to remove sampling variance from performance evaluation. A lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results.

@@

Evaluating imputation methodology requires a missing data problem to be solved. After establishing a comparative truth from a data-generating mechanism, some form of missingness therefore has to be induced. Often, however, reports of simulation studies remain vague about the actual missingness conditions under investigation and, even worse, some authors only report something like:
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness mechanisms and missingness patterns. Missingness mechanisms describe the relationship between missing entries and observed data values and are generally categorized into MCAR, MAR and MNAR mechanisms \citep[][see Table \ref{table:mech}]{rubi76}. [TODO: maybe remove mechanisms here?] Missingness patterns concern the location of missing entries across incomplete data \citep[][p. 8]{litt20}. Under this definition, there is a row-wise element to the missingness pattern describing which variables are jointly observed, and a column-wise element encompassing the amount of missingness in the data.

\begin{table}[tb]
\begin{center}
\caption{Missingness mechanisms. [TODO: add PMAR?]}
\label{table:mech}
\begin{tabular}{lll}
\hline
      & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & Missing completely at random: The probability to be missing is the same for all \\
      & cases. In other words, the missing values are missing at random and the observed \\
      & values are observed at random. \\
MAR   & Missing at random: The probability to be missing is the same within groups of \\
      & cases defined by the observed data only. The essence is that observed relations in \\
      & the data inform the missingness, such that MCAR may be assumed within the \\ 
      & observed groups. \\
MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
      & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
      & inferred from the observed data. This is considered non-ignorable missingness \\
      & \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

Even the terminology on missingness generation can be confusing. Does a missingness proportion of 50\% mean that half of the entries in an incomplete dataset are missing, or that half of the rows have at least one missing entry? In this paper, we will henceforth refer to the latter as the proportion of incomplete cases, and keep the term `missingness proportion' restricted to the variable-by-variable interpretation. The distinction between the two concepts, however, is not always clear in the literature, and convoluting the terms may lead to incorrect generalizations because they rarely mean the same (e.g., a proportion of 50\% incomplete cases in bivariate data could translate to a missingness proportion of 25\% in both variables, or one completely observed variable and one with 50\% missingness). Moreover, the value of the actual missingness proportion may be diffusing too. Some studies use only 10\% missingness whereas other studies push the limits to additionally investigate the performance under missingness proportions of at least 50\%. The inconsistent display of simulation conditions may impact the objectivity of meta-evaluations over imputation methods, as one method's performance may appear to be favorable because of less stringent simulation conditions. This ultimately may lead to statisticians recommending a less efficient method to applied researchers, thereby limiting the efficiency of the imputation approach and unnecessarily lowering the statistical power.

@@

Another aspect of missingness patterns that may misguide evaluations is the complexity of the patterns across variables. It is hardly reasonable to imagine empirical data with only one incomplete variable, yet some simulation studies rely on univariate missingness patterns anyways. Extrapolating simulation results from a specific missing data pattern to more intricate empirical missingness patterns could lead to sub-optimal advice in practice. For example, if a simulator induces missingness in the outcome variable of their analysis model exclusively, they may conclude that list-wise deletion outperforms their imputation method(s). Such a conclusion would, unfortunately, only translate to data that adhere to the same special case while biasing inferences in other cases \citep[][\S 2.7]{buur18}. Another example of a missingness pattern that may inadvertently impose assumptions on generalizations is a monotone pattern \citep[i.e. a pattern with uniformly increasing missingness proportions along variables;][]{litt20}. If a simulator investigates iterative imputation methods under monotone missingness, they might believe that the evaluated imputation algorithms converge instantly, while these methods actually require iteration to produce valid results in all other situations. A final example is multivariate missingness which is generated using step-wise univariate missingness induction. The resulting incomplete data may then not have the desired statistical properties \citep[e.g. skewness;][]{ampute}. This complicates any definitive conclusions about imputation method performance in relation to the data-generating mechanism.

% is the   [Other problems with missingness patterns: monotone/univariate vs multivariate, sporadic vs systematic, and univariately generated missing data may not have the expected multivariate patterns.] For example, iterative imputation algorithms converge instantly if there is only one variable with missingness or when the pattern is monotone, but require iteration in all other situations. And very specific missing data patterns may misguide which imputation method prevails in performance: some ad hoc methods are known to yield valid inferences under certain restricted conditions \citep[e.g., list-wise deletion may outperform imputation methods if there is missingness in the outcome variable of the analysis model exclusively][]{buur18}, while biasing inferences in other cases. 
% [Missingness mechanisms paragraphs: 1) definitions of mechanisms (now in table), 2) that MCAR is often ignored, while it's the minimum requirement and reference, and sometimes realistic in practice, 3) that MAR can be generated \textit{spuriously} if the correlation between variables is low,  4) that MNAR is often ignored, while it may be the most realistic in practice, and 5) finally add something about types of M(N)AR (ref Rianne's Dance paper).]
% MCAR missingness: The scenario where the missing values are missing at random and the observed values are observed at random. 
%  the validity of assumptions become increasingly important when the missingness increases.

Missingness mechanisms are usually assumed to be random (MAR; see Table \ref{table:mech} for definitions). However, the missingness mechanisms that are induced in simulation studies do not necessarily match any true missingness generating mechanisms in incomplete empirical data, nor the mechanisms that are typically assumed when imputing such data. On the one hand, there is MCAR missingness, which may be unlikely as the true missingness generating mechanism, but would yield insightful evaluations in a simulation setting. On the other hand, there is MNAR, which is arguably a quite reasonable source of many real-world missingness scenarios but is generally ignored except for those evaluations that are specifically targeted at non-ignorable applications. A disconnect between induced missingness, real missingness, and assumed missingness may result in simulation studies that are not as informative as they could be.

@@

Although MAR missingness is often considered as a simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to emulate a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is random. In fact, this would amount to one of the special cases under which complete case analysis would be more efficient than multiple imputation \citep[see e.g.][\S 2.7]{buur18}: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.

@@

The evaluation criteria used to assess imputation performance vary from one simulator to another. This is not surprising as people from different fields could have a different focus on the problem at hand. There are, however, some overarching issues with assessing imputation method performance. In the first place, diagnostic evaluation of the generated imputations is often left out of simulation study results. Problems with the imputation-generating process (e.g., an iterative imputation algorithm) may then be overlooked. In the second place, there are pitfalls in the evaluation of the estimates that are obtained by fitting the analysis model (i.e. the complete data model used to estimate the comparative truth). The choice of performance measures might [dictate (TODO: re-read and find better word!)] the results of the simulation study.

Many contemporary imputation techniques rely on iterative algorithms, such as the Gibbs sampler, to generate imputations. As with any iterative algorithm--but especially with imputation algorithms that are critically considered to be possibly incompatible Gibbs samplers \citep[PIGS,][]{li2012imputing}--algorithmic convergence should be carefully evaluated. Unfortunately, there is no universal quantitative method to diagnose non-convergence in iterative imputation algorithms \citep{zhu15, ober21} and the alternative \citep[visual inspection of the imputation algorithm;][]{buur18} is neither efficient nor failproof. As a result, imputation algorithms may be terminated before reaching a stable state, which could yield sub-optimal imputations and under-estimated performance of the method. Problems with producing stable imputations is not an exclusive quality iterative imputation algorithms. Any method may run into failures of the imputation-generating process and subsequently lack results (e.g. due to over-parameterization errors).

Omitting further evaluation of the imputation-generating process may yield sub-optimal imputations. The evaluation of simulated results relies on the relation between missing data models, imputation models, and analysis models. When an imputation model is able to capture the essence of the true non-response mechanism relative to the analysis model, the models are said to be congenial \citep{meng94}. The congeniality of all imputation models should be assessed, but methods to assess the suitability of imputation models and diagnose misfit often rely on visual inspection of the imputations \citep[see e.g.][]{abayomi2008diagnostics, bond16}. This makes its assessment an infeasible endeavor in many simulation studies. Moreover, the performance of imputation procedures on distributional properties is often ignored too. Even though the estimates on the analysis level may be justified, some methods can yield imputations that may seem completely invalid to applied researchers. For example, one could very accurately estimate average human height by filling in negative values and values that are unrealistically large. While the obtained inference could still be valid under such imputations, the plausibility of the imputed values given the observed data should be under scrutiny.

The second source of problems with evaluating simulation outcomes is the choice of performance measures. Performance measures should express how well the obtained estimates approximate the comparative truth, but not every metric is suited for the evaluation of imputation methodology. For example, using measures of predictive accuracy or the RMSE (root mean squared error) as the evaluation criterion may increase the rate of false positives \citep[][\S 2.6]{buur18}. And, under certain simulation conditions, an otherwise useful metric such as relative bias can distort the interpretation of results (i.e., a negligible absolute bias may yield an infinite relative bias if the true value of the estimand is zero). The usefulness of evaluation criteria may thus depend on the simulation goal and estimand.

If the goal is inference--not prediction--then the uncertainty about estimates needs to be properly quantified. To capture this uncertainty, the standard errors of the estimates should be correctly calculated, which requires multiple imputation. The aim of multiple imputation is not to reproduce the data, but to allow for obtaining valid inference given that the data are incomplete. This means that, given the framework provided by \citet{rubi87}, statistical properties such as bias, confidence intervals, and the coverage rate of the confidence intervals should be studied. After all, the 95\% confidence interval should contain the `true' value at least 95 out of 100 times \citep[][p. 591]{neym34}.

The choice of performance measures may inadvertently distort statistical properties of the imputed data. Developers often only inquire about the `accuracy' (i.e. how well can the method reproduce the original data). Merely focusing on the discrepancy between the true and imputed data would open the door for invalid inferences.

% The evaluation criteria may depend on the estimand. When descriptive statistics are the goal and when statistical inference would not be of interest, bias of the estimates would still apply, but standard errors are generally ignored.

% Some simulators may focus on retrieving the unobserved values, while others study the performance in the context of obtaining efficient predictions or classification.  
% For many outcome-driven imputation tasks a single valid imputation may perhaps suffice.

@@

\begin{table}[tb]
\begin{center}
\caption{Steps to consider in imputation simulation studies.}
\label{table:steps}
\begin{tabular}{lll}
\hline
                        & Simulation aspects under consideration \\
\hline  
1. Set scope            & Aim(s), missing data method(s) to evaluate, simulation design \\
2. Obtain truth         & Data-generating mechanism(s), estimand(s), sampling variance \\
3. Induce missingness   & Missingness mechanism(s), missing data pattern(s) \\
4. Apply methods        & Imputation method(s), analytic method(s) \\
5. Evaluate imputations & Algorithmic convergence, distributional characteristics, imputation \\
                        & model fit \\
6. Evaluate performance & Performance measures, qualitative judgment \\
7. Report               & Text, visualization, checklist, simulation script \\
\hline
\end{tabular}
\end{center}
% \footnotesize{Note. Multiple aspects may be varied across simulation conditions (e.g., using a factorial design with several data-generating mechanisms and missingness proportions).}
\end{table}

@@

Before setting up a simulation study, the simulator should clearly define the scope of their evaluations. A simulation study aimed at imputation method comparison will typically have a different design than one aimed at establishing the inferential validity of a single (novel) method. The choice of imputation method(s) under evaluation is naturally intertwined with the simulation aims and setup.

@@

Some general choices in the simulation setup to consider at this stage are the simulation design and the number of simulation repetitions. The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There are, for example, known interaction effects between missingness mechanisms and missingness patterns (the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions). If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons. The number of simulation repetitions may be informed by the required level of precision in the simulation study \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error][]{morr18}.

@@

The simulator should choose their data-generating mechanism(s) in line with the study scope. Model-based data-generating mechanisms have advantages in flexibility and precision, since data are generated from a known statistical model and the true theoretical parameters can be derived. A design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest.
